Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-25 VMCAI-revised-version/final.tex
up

\begin{proposition}\label{indec WSTS with dcs}
{\sc Resilience} is undecidable for reset-VASS, hence it is also undecidable for $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility and with 
$\Safe=\mathop{\downarrow} \Safe$.
\end{proposition}

\begin{proof}
First, note that {\sc Resilience} is undecidable for a Minsky machine $M$ with at least $3$ counters when
$\Safe = \mathop{\downarrow} \Safe$,
since it is possible to guarantee that state $q_f(0,0,0)$ is the only element 
% of $Q_{M} \times \mathop{\downarrow} (1,1,0)$
potentially reachable from 
$\Bad = Q_{M} \times \mathop{\uparrow} (0,0,1)$ that is not already inside $\Bad$.
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN,dufourd1998reset} in such a way that any sequence 
% from  a state $p({\bf u})$
% to
reaching
 a state 
%
$ q({\bf 0})$ corresponds to one that reaches 
$q_f(0,0,0)$
 in the Minsky machine and vice versa.
 Moreover, the reset-VASS is able to emulate zero tests on the counters, hence it is able to guarrantee that $ q({\bf 0})$ is its only $\Safe$ state reachable from a $\Bad$ state as well.
 By combining both constructions, which can be seen in more details in Appendix~\ref{minsk},
 we obtain that 
 {\sc Resilience} is undecidable for reset-VASS when $\Safe = \downarrow \Safe$, and hence for  $\post$-effective, $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed \end{proof}

@@

We now consider the case $\Safe = \mathop{\downarrow} \Safe$.
%
Unfortunately {\sc State-resilience} is undecidable in this case. As for {\sc Resilience} in WSTS when $\Safe = \mathop{\downarrow} \Safe$, this stems from undeciability of the corresponding problem in Minsky machines.
Indeed, {\sc State-resilience} is undecidable for a Minsky machine $M$ with at least $3$ counters when
$\Safe$ is downward-closed, since it is possible to guarantee that state $q_f(0,0,0)$ is the only $\Safe$ element potentially reachable from $\Bad$.
 As seen in 
Proposition~\ref{indec WSTS with dcs},
%\alain{la Proposition~\ref{indec WSTS with dcs} n'apparait que dans les annéexes...}
%\mathieu{Elle apparaît dans le corps du texte aussi mais j'avais utilisé deux fois le même label ce qui est corrigé maintenant.}
 the sequences of $M$ reaching a $\Safe$ state from a $\Bad$ one can be simulated in a reset-VASS
by sequences reaching a $\Safe'$ state from a $\Bad'$ one,
 and vice versa, hence the following result. 
% \alain{trop rapide}
%\mathieu{tjr trop rapide ?}
%\alain{ça va}

\begin{proposition}\label{srp down up}
{\sc State-resilience} is undecidable for reset-VASS, hence also for
%$\post$-ideal-effective ???? and
$\post$-effective,
 $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility
when
$\Safe=\mathop{\downarrow} \Safe$.
\end{proposition}

@@

\begin{proof}
Consider a given $k \in \N$,
 $\emptyset \neq \Bad \subseteq \N^d$ is upward-closed and $\Safe \subseteq \N^d$ is downward-closed, and
consider a given VAS $V$.
Let us call $c_{\max}$ the maximal absolute value of a constant appearing in a coordinate of a transition of $V$.
The set $\Bad \neq \emptyset$ admits a finite basis $B_\Bad$.
Consider the vector $\textbf{v}_{\Bad}$ obtained by summing all members of the basis of $\Bad$ and then consider the vector
$\textbf{u}_k = \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, c_{\max}, \ldots, c_{\max})$.

@@

\section{Reset-VASS and Minsky-machines}\label{minsk}

We now delve into more details into the undecidability results 
upon which rely
Proposition~\ref{indec WSTS with dcs}
and
Proposition~\ref{srp down up},
namely, that
{\sc Resilience} and {\sc State-resilience} when $\Safe = \mathop{\downarrow}~\Safe$
are undecidable for Minsky machines
and 
for reset-VASS.

The problems are undecidable for a Minsky machine with at least $3$ counters when
$\Safe$ is downward-closed, since it is possible to guarantee that state $q_f(0,0,0)$ is the only $\Safe$ element potentially reachable from $\Bad$. We then consider a reset-VASS simulating the Minsky machine. We then use Reset-VASS to simulate the Minsky machines~\cite{araki1976PN,dufourd1998reset}, albeit in a weak way. 
% The simulation however is 
Since by a particular choice of $\Safe$ and $\Bad$ we are only interested in the reachability of $q_f(0,0,0)$ in the Minsky machine, 
this weak simulation will be enough to obtain undecidability of both problems.

\begin{proposition}\label{both minsk indec}
{\sc Resilience} and {\sc State-resilience} when $\Safe = \mathop{\downarrow}~\Safe$ are undecidable for reset-VASS, hence also undecidable for $\post$-effective, $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility.
\end{proposition}

\begin{proof}
\alain{preuve pas claire, à réécrire}
% \alain{les reset-VASS sont-ils effective $\post^*$-basis ?}
% \mathieu{non, sinon on aurait décidabilité pour la state résilience}
%This stems from the fact that it is undecidable for Minsky machines with more than one counter,
%which can be simulated by WSTS.
First, let us show that {\sc Resilience} is undecidable for $3$-counter machines.
%	This is due to the undecidability of termination for $2$-counter Minsky machines.
From an $2$-counter machine $M=(Q, q_f,...)$ where $Q$ is the set of control-states and $q_f \in Q$ is the final control-state, one constructs an associated $3$-counter machine $M'=(Q', q'_f,...)$ with a final control-state $q_f' \in Q'$
such that $M$ terminates on all initial states iff $q_f'(0,0,0)$ is reachable (in $M'$) from any state in $Q' \times \mathop{\uparrow} (0,0,1)$. 
%			with at least $1$ on its third counter. 
We build $M'$ to simulate $M$ until it reaches $q_f$, then $M'$ decreases its first two counters until they reach $0$, then, and only then, finally $M'$ decreases the third counter until it reaches $0$.
%
Remark that $Q' \times \{ (m,0,0), (0,n,0),(m,n,0) \mid m,n \in \N\}$ is not reachable from any state in $Q' \times \mathop{\uparrow}(0,0,1)$.
%
Based on this construction, the downward-closed set 
$Q' \times (\N^3 \setminus \mathop{\uparrow}(0,0,1))$ is reachable (in $M'$)
% \alain{review 3: - p. 10: why not choose $\mathop{\downarrow}{(0,0,0)}$ instead of $\mathop{\downarrow}{(1,1,0)}$?}
from any state in $Q' \times \mathop{\uparrow}(0,0,1)$ iff $M$ terminates on all initial states; this last problem is undecidable~\cite{Min61}. 
Note that by chosing $s_0 \in Q' \times \mathop{\uparrow} (0,0,1)$,
deciding whether the downward-closed set 
$Q' \times (\N^3 \setminus \mathop{\uparrow}(0,0,1))$ is reachable
from every state in $\post^*(s_0) \cup Q' \times \mathop{\uparrow} (0,0,1)$ 
is again equivalent to asking whether $M$ terminates on every initial state.
Hence {\sc Resilience} and {\sc State-resilience} are undecidable for Minsky machines.

% \alain{écrire qqchose ici}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN,dufourd1998reset}. 
% \alain{\cite{araki1976PN} simule-t-il les Minsky machines et comment ?}
%Reset-VASS extend the basic VASS model with special “reset
%transitions” that set to $0$ some coordinates in the vector. 
%			Let us recall their definition here.
%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}.
The simulation operates by building a reset-VASS in which
there exists
 a sequence from 
an initial state %$(s,x_1 \ldots x_n)$
to
a state $ q({\bf 0})$ if and only if
there exists
a sequence
 in the Minsky machine
 from a corresponding initial state
 to
 the state $q_f(0,0,0)$
  (from Theorem~$5$ in \cite{araki1976PN}).
  Moreover, the reset-VASS is able to emulate zero tests on the counters, hence it is able to guarrantee that $ q({\bf 0})$ is its only $\Safe$ state reachable from a $\Bad$ state as well.
Hence {\sc Resilience} (resp. {\sc State resilience}) holds in the Minsky machine $M'$ iff 
{\sc Resilience} (resp. {\sc State resilience}) holds in the reset-VASS simulating $M'$.
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} 
and  {\sc State-resilience} are undecidable for reset-VASS and hence for  $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed
%\alain{déjà dit avant en mieux omega2 et strong comp et $\post$-effective mais reset-VASS apportent effective $\pred$-basis et ideally-effective....}
\end{proof}

>>>>

\begin{proposition}\label{indec WSTS with dcs}
{\sc Resilience} is undecidable for reset-VASS when $\Safe=\mathop{\downarrow} \Safe$, hence it is also undecidable for $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility.
\end{proposition}

\begin{proof}
\nocite{araki1976PN,dufourd1998reset}
{\sc Resilience} when $\Safe= \{ q_f({\bf 0}) \}$ for some state $q_f$ is equivalent to
deciding structural termination of reset-VASS, which is undecidable.
\qed \end{proof}
\mathieu{GOTO}

@@

We now consider the case $\Safe = \mathop{\downarrow} \Safe$.
%
Unfortunately {\sc State-resilience} is undecidable in this case. As for {\sc Resilience} in WSTS when $\Safe = \mathop{\downarrow} \Safe$, this stems from undeciability of the 
structural termination of reset-VASS.
\mathieu{GOTO}

\begin{proposition}\label{srp down up}
{\sc State-resilience} is undecidable for reset-VASS when
$\Safe=\mathop{\downarrow} \Safe$, hence also for
$\post$-effective,
 $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility.
 \end{proposition}

@@

\begin{proof}
Consider a given $k \in \N$, $\Safe \subseteq \N^d$ is downward-closed
 such that its upward-closed complement $\Bad \subseteq \N^d$ is nonempty, and
consider a given VAS $V$. 
Let us call $c_{\max}$ the maximal absolute value of a constant appearing in a coordinate of a transition of $V$.
The set $\Bad \neq \emptyset$ admits a finite basis $B_\Bad$.
Consider the vector $\textbf{v}_{\Bad}$ obtained by summing all members of the basis of $\Bad$ and then consider the vector
$\textbf{u}_k = \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, c_{\max}, \ldots, c_{\max})$.

@@
